---
title: "STA 322 Project 2"
author: "Alayna Binder & Katherine Zhong"
date: "2025-12-05"
output: 
  bookdown::pdf_document2:
    number_sections: true
    toc: false
---

# Experimental Design

We conducted a full $2^4$ factorial experiment using all sixteen combinations of the four helicopter design factors A, B, C, and D. A factorial structure provides the most information per trial because it allows us to estimate main effects and interactions simultaneously, and it avoids the inefficiency of changing only one factor at a time. For each design combination, we collected four replicate flights. This level of replication follows the project guidance and provides reasonable accuracy for our effect estimates while keeping the workload manageable.

To prevent ordering effects from influencing the results, we randomized the flight order before collecting any data. Randomizing the sequence of the sixteen design combinations ensured that no configuration was consistently flown earlier or later in the session. After determining the order of combinations, we also randomized the assignment of each flight to a thrower. Both throwers contributed flights across all sixteen combinations, creating a balanced structure in which every design was flown by both people. Throughout data collection, each thrower made a conscious effort to drop the helicopter the same way every time by maintaining a consistent release height, orientation, and motion. We tried to minimize variability wherever possible so that observed differences in flight time would reflect the design factors rather than differences in technique.

We blocked on thrower because the person releasing the helicopter can influence flight time independently of the design. Blocking allows us to treat the thrower as a nuisance factor that is controlled for rather than ignored. Differences in release style, timing, or consistency can introduce extra variability if they are not accounted for, and including thrower as a block directly models these differences. Since both throwers completed flights for every design combination, the block structure is balanced and contributes to more precise estimates of the factor effects.

All flights were completed in one session. Although repeated measurements within a session may raise concerns about fatigue or small changes in technique, the total number of throws per person was modest. Any minor shifts in performance would be spread throughout the experiment because of the randomization, rather than concentrated in specific design conditions. For this reason, and because we observed no systematic drift in the flight times, we assumed that any within-session variation was minimal.

```{r echo = FALSE, message = FALSE, warning = FALSE}
#| label: load-packages

library(tidyverse)
library(broom)
library(knitr)
library(kableExtra)
```

```{r echo = FALSE, message = FALSE, warning = FALSE}
#| label: load-data

data <- read_csv("STA322-project2-helicopter - Sheet1.csv")
```

```{r echo = FALSE, message = FALSE, warning = FALSE}
#| label: data-wrangling

df <- data |>
  mutate(
    # treat a–d as 0/1 indicators for the four design factors
    across(a:d, as.integer),
    
    # factors for people
    thrower = factor(thrower),
    maker = factor(maker),
    
    # combination as a factor, ordered by how it appears
    combination = factor(combination, levels = unique(combination))
  )

# summary(df$time)
```

# Analysis

To identify which design factors influence flight time, we first examined the distributions of flight times across the two levels of each factor. The main effect boxplots show a clear upward shift for factor A, with level 1 producing noticeably longer flights than level 0. Factor D shows the opposite pattern, with shorter flights at level 1. In contrast, factors B and C display only small differences between levels, and the variation within each level is larger than any apparent shift between them.

```{r echo = FALSE, message = FALSE, warning = FALSE, fig.width = 7, fig.height = 3, fig.pos = "H"}
#| label: main-effects

main_effect_summaries <- df |>
  pivot_longer(a:d, names_to = "factor", values_to = "level") |>
  group_by(factor, level) |>
  summarise(
    mean_time = mean(time),
    sd_time = sd(time),
    n = n(),
    .groups = "drop"
  )

# main_effect_summaries

df |>
  pivot_longer(a:d, names_to = "factor", values_to = "level") |>
  mutate(level = factor(level)) |>
  ggplot(aes(x = level, y = time)) +
  geom_boxplot() +
  facet_wrap(~ factor, nrow = 1) +
  labs(
    title = "Main Effect Plots for Factors A-D",
    x = "Level",
    y = "Flight time (seconds)"
  )
```

We then formalized these patterns using a linear model that included all four design factors while blocking on thrower. The estimated effects and confidence intervals are shown in Table 1. Factor A had a positive and statistically significant effect, increasing the mean flight time by about 0.22 seconds. Factor D had a negative and significant effect, decreasing the mean flight time by about 0.14 seconds. The confidence intervals for both effects were entirely away from zero. Factors B and C had small estimated effects with confidence intervals that included zero, indicating little evidence that they meaningfully affect flight time.

```{r echo = FALSE, message = FALSE, warning = FALSE, table.env = FALSE}
#| label: Q1

fit_q1 <- lm(time ~ a + b + c + d + thrower, data = df)

# summary(fit_q1)

q1_table <- tidy(fit_q1, conf.int = TRUE) |>
  select(term, estimate, conf.low, conf.high, std.error, statistic, p.value)

q1_table |>
  kable(digits = 3, caption = "Main Effects Model Results") |>
  kable_classic(full_width = FALSE)
```

Both the exploratory plots and the fitted model point to the same conclusion. Factors A and D are the most influential design components. Factor A corresponds to rotor length, and the model indicates that longer rotors tend to increase flight time. Factor D corresponds to whether a paper clip is attached, and the negative effect estimate suggests that adding a paper clip shortens the flight. In contrast, the variables associated with leg length (factor B) and leg width (factor C) show minimal impact on flight duration under this experimental setup.

